Verification of Stencil Computations through Spatial Specifications

### 2016-07-06 paper.tex
edits to section 3 and 4

\begin{definition}%[Schematic]
The partial function $\textsf{schematic}$ maps
source indexing terms $(e_1, .., e_n)$
to $n$-dimensional index schemes, defined:
%
\begin{align*}
\mathsf{schematic}(e_1, \ldots, e_n) & =
[\downarrow\!e_1 \ldots \downarrow\!e_n ]
\quad \textit{iff} \;\; \forall i . \downarrow\!e_i \neq \bot
\end{align*}
%
The $\downarrow$ operator is partial\footnote{We represent
  partiality in our signatures with the set $A + \bot$.} 
 $\downarrow : \textit{expr} \rightarrow (\domainVal + \bot)$
mapping individal indexing expressions to 
an indexing scheme component, defined:
%%
\begin{align*}
\downarrow\!e
 & =  \begin{cases}
a & e \equiv i + a \; \wedge \; \mathsf{neigh}(i + a) \\
-a & e \equiv i - a \; \wedge \; \mathsf{neigh}(i - a) \\
\infty & \textit{$\mathsf{IV}(e) = \emptyset$} \\
\bot   & \textit{otherwise}
\end{cases}
\end{align*}
%%
where $\mathsf{IV}(e)$ is the set of free induction variables
appearing in an expression, and $\neigh{}$ classifies expressions which
are constant offsets from an induction variable
(Defn.~\ref{def:neighbour}). In the third case,
if the expression is not a neighbourhood index but is not
a relative index either (\ie{}, not defined in terms
of an induction variable) then it is mapped to $\infty$, otherwise
$\downarrow\!e$ is undefined. The $\mathsf{schematic}$ function
is undefined if $\downarrow$ of any of its components $e_i$ is
undefined.

@@

The intuition behind $\otimes$ is that it takes
all possible pairs of indexing schemes, treating them as the
lower and upper bounds of an $n$-dimensional rectangle from which the
remaining vertices of the rectangle are generated (like a bounding
box) \eg{}:
%%
\begin{align*}
\{\vtwohs{1}{2}, \vtwohs{5}{6}\,\} \otimes \{\vtwohs{3}{4}\} =
\{\vtwohs{1}{2}, \vtwohs{3}{2}, \vtwohs{1}{4}, \vtwohs{3}{4}, 
   \vtwohs{3}{6}, \vtwohs{5}{4}, \vtwohs{5}{6}\,\}
\end{align*}
%%
The tensor takes a Cartesian product of the two models,
giving all pairs of schemes which are combined using the
\emph{pairwise permutation} written $\bowtie$ (defined later).
Pairwise permutation generates all possible interleavings of two vectors
(\ie{}, all possible non-deterministic choices between $u_i$ and
$v_i$ in each dimension). Some care is then taken over
how this interacts with $\infty$: pairwise permtuation
corresponds to non-determinstic choice, in each dimension $i$,
between two values $u_i$ and $v_i$, but if either
is $\infty$ then the resultying pairwise permutation is
filtered out if the dimension $i$ has \emph{any other scheme
in the model where $u_i \neq \infty$}, \eg{}:
\begin{align*}
\{\vtwoh{1}{\infty}\} \otimes \{\vtwoh{3}{4}\} =
\{\vtwoh{1}{4}, \vtwoh{3}{4}\}
\end{align*}
%
where $\vtwoh{1}{\infty}$ and $\vtwoh{3}{\infty}$ have been filtered.
%With respect to the notion of unification, the
%
%
% combines non-$\infty$ constraints
%of the unifiers in different dimensions into less accepting unifiers.
%This means an access pattern generalises with this scheme would require
%both models to be satisfied. Use of $dims$ gives the dimensions that
%constitutent models talk about and is used in the product to eliminate
%general unifier appearing in these dimensions. Allowing general unifier
%in these dimensions would prevent both models to be satisfied
%simultaneously.
%  u in M * N   u < w
% mplies that
%  u_1 in M  u1 < w
%  u_2 in M  u2 < w
%
The notion of which dimensioons $i$ have that \emph{any scheme in the model
has $u_i \neq \infty$} is represented by the \emph{constrained
dimensions} intermediate function:
%
\begin{definition}[Constrained dimensions]
Given an $n$-dimensional model $M$, the \emph{constrained dimensions}
of that model is a subset of $\{1, \ldots, n\}$ of
the dimensions for which at least one scheme $u \in M$ has
a non-$\infty$ value in that dimension. That is:
%
\begin{align*}
\effdims{M}{n} =
\bigcup_{u \in M} \{i \mid i \in \{1,\ldots,n\} \, \wedge \, u_i \neq
  \infty\}
\end{align*}
For example, $\effdims{\{\vtwoh{0}{\infty},\vtwoh{\infty}{\infty}\}}{2}
 = \{1\}$.
\end{definition}

\begin{definition}The tensor $\otimes_n$ of $n$-dimensional
models is defined:
\begin{alignat*}{2}
& M \otimes_n N =
  \Bigg\{x \; \Bigg| \;
    \parbox{6cm}{$(u, v) \in M \times N $ \\[0.1em]
                 $\;\; \wedge \;\, j \in (\effdims{M}{n} \cup
                   \effdims{N}{n})$\\[0.1em]
                  $\;\; \wedge \;\, i \in \{1, \ldots, 2^n\} \, \wedge
                  \, x = (u \bowtie v)_i$ \\[0.1em]
                   $\;\; \wedge \;\, x_j \neq \infty$
                  } \Bigg\}
\end{alignat*}
The first guard takes the Cartesian product of
the two models and quantifies over all pairings $(u, v)$.
The second guard says that $j$ ranges of the constrained
dimensions for $M$ unioned with the constrained dimensions for $N$.
That is, every $j$ corresponds to a dimension in which
either $M$ or $N$ has a scheme $u$ where $u_j \neq \infty$,
and thus is constrained.

The $\bowtie$ \emph{pairwise permutation}
constructs a $2^n \times n$ matrix ($2^n$-vector of
$n$-vectors), which is indexed by $i$ above
and then $j$. Using this, the $\otimes$ operation constructs the set
of schemes which are the
pairwise permutations of every pair $(u, v)$ of schemes in the Cartesian
product of $M \times N$. These permuted schemes are only those
which have non-$\infty$ components in the constrained dimensions
given by each model, given by the guard $x_j \neq \infty$.

The pairwise permutation $u \bowtie v$ takes two vectors and builds the matrix
where the rows are all possible $n$-vectors generated by
non-deterministic picking each $i^{th}$ entry from either the
$i^{th}$ entry in $u$ or the $i^{th}$ entry in $j$. For example:
%
\begin{equation*}
\vthreeh{0}{1}{2} \bowtie \vthreeh{3}{4}{5} =
\left[
\setlength{\arraycolsep}{0.42em}
\begin{array}{cccccccc}
0 & 0 & 0 & 0 & 3 & 3 & 3 & 3 \\
1 & 1 & 4 & 4 & 1 & 1 & 4 & 4 \\
2 & 5 & 2 & 5 & 2 & 5 & 2 & 5
\end{array}
\right]
% \\
%\setlength{\arraycolsep}{0.5em}
%\left[\begin{array}{ccc}
%0 & 1 & 2 \\
%0 & 1 & 5 \\
%0 & 4 & 2 \\
%0 & 4 & 5 \\
%3 & 1 & 2 \\
%3 & 1 & 5 \\
%3 & 4 & 2 \\
%3 & 4 & 5
%\end{array}
%\right]
\end{equation*}
%
The $2^n$ unique choices for $\bowtie$ on $n$-vectors
corresponds to taking all bit-strings of length $n$ and
selecting from $u$ for $1$ and $v$ for $0$. The pairwise permutation
matrix is thus defined in terms of the logical
matrix $b$ where $b_{i,j}$ is the $j$-th bit of the integer
$i - 1$ as:
%
\begin{align*}
(u \bowtie v)_{i,j} & = b_{i,j} u_j + \neg b_{i,j} v_j
\end{align*}
%
\end{definition}

@@

\begin{definition} The \textsf{Mult}
and \textsf{Approx} are parametric labelled variant (sum) types
with injection function given by their definition:
%
\begin{align*}
\textsf{Mult} \, a & = \textsf{mult} \, a \,\mid\, \textsf{only} \, a \\
\textsf{Approx} \, a & = \textsf{exact} \, a \,\mid\, \textsf{lower} \, a \,\mid\,
\textsf{upper} \, a \,\mid\, \textsf{both} \, a \, a
\end{align*}
\ie{}, $\mathsf{lower}$ is an injection $\mathsf{lower} : a \rightarrow \mathsf{Approx} \, a$, and so
on. These are used for labelled models, and other components later in
Section~\ref{sec:analysis} with additional information.
\label{def:mult-and-approx}
\end{definition}
%

\begin{figure}
\input{spatial-model.tex}
\caption{\textit{spec} model,
$\interp{-}_n : \textit{spec} \rightarrow \textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))$}
\label{fig:spatial-model}
\end{figure}

@@

Given the model defined in this section we can
consider soundness for the equational theory of
specifications and the notion of spatial approximation. The proofs are given in our supplement.

@@

Let $\leq$ be an ordering on
$\textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))$
and $\leq'$ be an ordering on
$\textsf{Approx}(\mathcal{P}(\domainVal{}^n))$ defined by the
following caluses
\begin{align*}
(M \leq' N \Rightarrow \textit{inj} \, M \leq \textit{inj} \, N) &
\qquad 
(\textsf{once} \, M \leq' \textsf{mult} \, M) \\
%%
%%
(M \subseteq N \Rightarrow \textit{inj}' M \leq' \textit{inj}' N)
& \;\;\; (\textsf{exact} M \leq  \textsf{lower} M)
\;\;\; (\textsf{lower} M \leq  \textsf{upper} M)
\end{align*}
%%
where $\textit{inj}$ ranges over the unary injections
of $\textsf{Approx}$ and $\textit{inj}'$ ranges over the injections of
$\textsf{Mult}$. Then, soundness of $<:$ is that:
%%
\[
S <: T
\; \Rightarrow \;
\interp{S}_n \leq \interp{T}_n
\]
\end{theorem}

@@

\noindent
We briefly sketch the algorithms for inferring and checking
specifications and for synthesising programs from specifications.
Checking and synthesis use the set-based semantic model of
Section~\ref{sec:semantics}.

\paragraph{Syntax functions}
The function \rhsExp{} maps statements to a
set of expressions which occur in right-hand positions (\ie{}, not the
target of an assignment). The function \var{} maps expressions to a
set of the variables used in right-hand positions.

\subsection{Inference\label{subsec:inference}}

We demonstrate the main steps of the inference procedure over the
following example program which computes the mean value
of a five-point stencil at each index of the input array:
%%
\begin{minted}{fortran}
do i = 1, (n-1)
   do j = 1, (m-1)
      x       = a(i-1, j) + a(i+1, j)
      y       = a(i, j-1) + a(i, j+1)
      b(i, j) = (a(i, j) + x + y) / 5.0
   end do
end do
\end{minted}
%%
\subsubsection{Step 1: Standard control and data-flow analyses}
\label{sec:inf-step1}

@@

For each assignment statement whose left-hand side is an array
subscript on neighbourhood indices, a finite map is computed which
maps array variables to a set of vectors representing array
subscripts.  This finite map contains all array subscript expressions
which flow to this statement. More formally, a function
$\textsf{analyse}$ is applied to each statement in a program with the
following clause:
%
\begin{align*}
\begin{array}{lr}
\textsf{analyse}(v(\overline{e_1}) = e_2)
 := \qquad\qquad & \textit{where} \; \neigh(\overline{e_1}) \wedge \arrayTy(v)  \\[0.3em]
\multicolumn{2}{l}{\qquad \bigcup\{v' \mapsto \{\textsf{schematic}(\bar{e})\} \mid v'(\bar{e}) \leftarrow \mathsf{flowsTo}(e_2),
  \arrayTy(v')\}}
\end{array}
\end{align*}
%
That is, we focus on assignments to an array subscript where each
indexing expression in $\bar{e}$ is a neighbourhood index.  For
all array subscripts that flow to the right-hand side of this
statement, a finite map is constructed, mapping each array variable
in this flow a set of indexing schemes for its subscripts, computed
with \textsf{schematic}.

For our example, the \textsf{analyse} function matches on
line 5, with the following set for $\textsf{flowsTo}(\texttt{a(i, j) + x +
  y})$:
%
\begin{align*}
\{\texttt{a(i-1, j)}, \texttt{a(i+1, j)}, \texttt{a(i, j-1)},
  \texttt{a(i, j+1)}, \texttt{a(i, j)}\}
\end{align*}
Subsequently the result of \textsf{analyse} on line 5 yields the map:
\begin{align*}
\texttt{a} \mapsto \{\vtwoh{-1}{0}, \vtwoh{1}{0},
          \vtwoh{0}{-1}, \vtwoh{0}{1}, \vtwoh{0}{0}\}
\end{align*}
%

>>>>

\begin{definition}%[Schematic]
\label{def:schematic}
The partial function $\textsf{schematic}$ maps
source indexing terms $(e_1, .., e_n)$
to $n$-dimensional index schemes, defined:
%
\begin{align*}
\mathsf{schematic}(e_1, \ldots, e_n) & =
[\downarrow\!e_1 \ldots \downarrow\!e_n ]
\quad \textit{iff} \;\; \forall i . \downarrow\!e_i \neq \bot
\end{align*}
%
The $\downarrow$ operator is partial\footnote{We represent
  partiality in our signatures with the set $A + \bot$.} 
 $\downarrow : \textit{expr} \rightarrow (\domainVal + \bot)$
mapping individal indexing expressions to 
an indexing scheme component, defined:
%%
\begin{align*}
\downarrow\!e
 & =  \begin{cases}
a & e \equiv i + a \; \wedge \; \mathsf{neigh}(i + a) \\
-a & e \equiv i - a \; \wedge \; \mathsf{neigh}(i - a) \\
\infty & \textit{$\mathsf{IV}(e) = \emptyset$} \\
\bot   & \textit{otherwise}
\end{cases}
\end{align*}
%%
where $\mathsf{IV}(e)$ is the set of free induction variables
appearing in an expression, and $\neigh{}$ classifies expressions which
are constant offsets from an induction variable
(Defn.~\ref{def:neighbour}). In the third case,
if the expression is not a neighbourhood index but is not
a relative index either (\ie{}, not defined in terms
of an induction variable) then it is mapped to $\infty$, otherwise
$\downarrow\!e$ is undefined. The $\mathsf{schematic}$ function
is undefined if $\downarrow$ of any of its components $e_i$ is
undefined.

@@

The intuition behind $\otimes$ is that it takes
all possible pairs of indexing schemes, treating them as the
lower and upper bounds of an $n$-dimensional rectangle from which the
remaining vertices of the rectangle are generated (like a bounding
box) \eg{}:
%%
\begin{align*}
\{\vtwohs{1}{2}, \vtwohs{5}{6}\,\} \otimes \{\vtwohs{3}{4}\} =
\{\vtwohs{1}{2}, \vtwohs{3}{2}, \vtwohs{1}{4}, \vtwohs{3}{4}, 
   \vtwohs{3}{6}, \vtwohs{5}{4}, \vtwohs{5}{6}\,\}
\end{align*}
%%
The tensor takes a Cartesian product of the two models,
giving all pairs of schemes which are combined using the
\emph{pairwise permutation} written $\bowtie$ (defined later).
Pairwise permutation generates all possible interleavings of two vectors
(\ie{}, all possible non-deterministic choices between $u_i$ and
$v_i$ in each dimension). Some care is then taken over
how this interacts with $\infty$: pairwise permtuation
corresponds to non-determinstic choice, in each dimension $i$,
between two values $u_i$ and $v_i$, but if either
is $\infty$ then the resulting pairwise permutation is
filtered out if the dimension $i$ has \emph{any other scheme
in the model where $u_i \neq \infty$}, \eg{}:
\begin{align*}
\{\vtwoh{1}{\infty}\} \otimes \{\vtwoh{3}{4}\} =
\{\vtwoh{1}{4}, \vtwoh{3}{4}\}
\end{align*}
%
where $\vtwoh{1}{\infty}$ and $\vtwoh{3}{\infty}$ have been filtered.
%With respect to the notion of unification, the
%
%
% combines non-$\infty$ constraints
%of the unifiers in different dimensions into less accepting unifiers.
%This means an access pattern generalises with this scheme would require
%both models to be satisfied. Use of $dims$ gives the dimensions that
%constitutent models talk about and is used in the product to eliminate
%general unifier appearing in these dimensions. Allowing general unifier
%in these dimensions would prevent both models to be satisfied
%simultaneously.
%  u in M * N   u < w
% mplies that
%  u_1 in M  u1 < w
%  u_2 in M  u2 < w
%
The notion of which dimensions $i$ have that \emph{any scheme in the model
has $u_i \neq \infty$} is represented by the \emph{constrained
dimensions} intermediate function:
%
\begin{definition}%[Constrained dimensions]
For an $n$-dimensional model $M$, its \emph{constrained dimensions}
is the set of dimension identifiers for which at least one scheme $u \in M$ has
a non-$\infty$ value in that dimension. That is:
%
\begin{align*}
\effdims{M}{n} =
\bigcup_{u \in M} \{i \mid i \in \{1,\ldots,n\} \, \wedge \, u_i \neq
  \infty\}
\end{align*}
For example, $\effdims{\{\vtwoh{0}{\infty},\vtwoh{\infty}{\infty}\}}{2}
 = \{1\}$.
\end{definition}

\begin{definition}The tensor $\otimes_n$ of $n$-dimensional
models is defined:
\begin{alignat*}{2}
& M \otimes_n N =
  \Bigg\{x \; \Bigg| \;
    \parbox{5.2cm}{$(u, v) \in M \times N $ \\[0.1em]
                 $\;\; \wedge \;\, j \in (\effdims{M}{n} \cup
                   \effdims{N}{n})$\\[0.1em]
                  $\;\; \wedge \;\, i \in \{1, \ldots, 2^n\} \, \wedge
                  \, x = (u \bowtie v)_i$ \\[0.1em]
                   $\;\; \wedge \;\, x_j \neq \infty$
                  } \Bigg\}
\end{alignat*}
The first guard takes the Cartesian product of
the models and quantifies over all pairings $(u, v)$.
The second guard binds $j$ which ranges over the constrained
dimensions for $M$ union with the constrained dimensions for $N$.
That is, every $j$ corresponds to a dimension in which
either $M$ or $N$ has a scheme $u$ where $u_j \neq \infty$,
and thus is constrained.

The \emph{pairwise permutation} $u \bowtie v$ 
builds a $2^n \times n$ matrix 
of all possible $n$-vectors generated by
non-deterministically picking for each $j^{th}$ entry either 
$u_j$ or $v_j$ \eg{}
%
\begin{equation*}
\vthreeh{0}{1}{2} \bowtie \vthreeh{3}{4}{5} =
\left[
\setlength{\arraycolsep}{0.42em}
\begin{array}{cccccccc}
0 & 0 & 0 & 0 & 3 & 3 & 3 & 3 \\
1 & 1 & 4 & 4 & 1 & 1 & 4 & 4 \\
2 & 5 & 2 & 5 & 2 & 5 & 2 & 5
\end{array}
\right]
% \\
%\setlength{\arraycolsep}{0.5em}
%\left[\begin{array}{ccc}
%0 & 1 & 2 \\
%0 & 1 & 5 \\
%0 & 4 & 2 \\
%0 & 4 & 5 \\
%3 & 1 & 2 \\
%3 & 1 & 5 \\
%3 & 4 & 2 \\
%3 & 4 & 5
%\end{array}
%\right]
\end{equation*}
%
The $2^n$ unique choices for $\bowtie$ on $n$-vectors
corresponds to taking all bit-strings of length $n$ and
selecting from $u$ for $1$ and $v$ for $0$. The pairwise permutation
matrix is thus defined in terms of the logical
matrix $b$ where $b_{i,j}$ is the $j$-th bit of the integer
$i - 1$ as:
%
\begin{align*}
(u \bowtie v)_{i,j} & = b_{i,j} u_j + \neg b_{i,j} v_j
\end{align*}
%
The permutated schemes $x = (u \bowtie v)_i$ added to the model 
$M \otimes N$ are only those which have do not have $\infty$ in 
the constrained dimensions given by each model ($x_j \neq
\infty$). This filters out all permutations which have
$\infty$ in a constrained dimension position. 
\end{definition}

@@

\begin{definition} \textsf{Mult}
and \textsf{Approx} are parametric labelled variant types
with injection function given by their definition:
%
\begin{align*}
\textsf{Mult} \, a & = \textsf{mult} \, a \,\mid\, \textsf{only} \, a \\
\textsf{Approx} \, a & = \textsf{exact} \, a \,\mid\, \textsf{lower} \, a \,\mid\,
\textsf{upper} \, a \,\mid\, \textsf{both} \, a \, a
\end{align*}
\ie{}, $\mathsf{lower}$ is an injection $\mathsf{lower} : a \rightarrow \mathsf{Approx} \, a$, and so
on. The $\textsf{Mult}$ type corresponds to the presence or absence of the
\term{readOnce} modifier as shown in Figure~\ref{fig:spatial-model}. 
The $\textsf{Approx}$ type corresponds to the presence or absence of
the spatial approximation modifier, with \textsf{exact} when there is
no such modifier and \textsf{lower} and \textsf{upper} for
\term{atLeast} and \term{atMost}. Later, it will be useful to model
pairs of lower and upper bounds which is provided by \textsf{both} 
(but which isn't used here). These are used for labelled models, and other components later in
Section~\ref{sec:analysis}.% with additional information.
\label{def:mult-and-approx}
\end{definition}

\begin{figure}
\input{spatial-model.tex}
\caption{\textit{spec} model,
$\interp{-}_n : \textit{spec} \rightarrow \textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))$}
\label{fig:spatial-model}
\vspace{-0.75em}
\end{figure}

@@

Given the above model, we can
consider soundness for the equational theory of
specifications and the notion of spatial approximation. The proofs are given in our supplement.

@@

Let $\leq$ be an ordering on
$\textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))$
and $\leq'$ be an ordering on
$\textsf{Approx}(\mathcal{P}(\domainVal{}^n))$ defined by the
following clauses:
\begin{align*}
\def\arraystretch{1.2}
\begin{array}{c}
\begin{array}{cc}
(M \leq' N \Rightarrow \textit{inj} \, M \leq \textit{inj} \, N) &
(\textsf{once} \, M \leq' \textsf{mult} \, M) \\
%%
%%
(M \subseteq N \Rightarrow \textit{inj}' M \leq' \textit{inj}' N)
& (\textsf{exact} M \leq  \textsf{lower} M)
\end{array} \\ 
(\textsf{lower} M \leq  \textsf{upper} M)
\end{array}
\end{align*}
%%
where $\textit{inj}$ ranges over the unary injections
of $\textsf{Approx}$ and $\textit{inj}'$ ranges over the injections of
$\textsf{Mult}$. Then, soundness of $<:$ is that:
%%
\[
S <: T
\; \Rightarrow \;
\interp{S}_n \leq \interp{T}_n
\]
\end{theorem}

@@

\noindent
We briefly sketch the algorithms for inferring and synthesising
specifications (Sect.~\ref{subsec:inference}), for checking code
against specifications (Sect.~\ref{subsec:checking}) and for
synthesising programs from specifications
(Sect.~\ref{subsec:synthesis}).  These utilise the semantic model of
Section~\ref{sec:semantics}.

\paragraph{Syntax helper functions}
The function \rhsExp{} maps source statements to a
set of expressions which occur in right-hand positions (\ie{}, not the
target of an assignment). The function \var{} maps expressions to a
set of the variables used in right-hand positions.

\subsection{Inference and specification synthesis\label{subsec:inference}}

We demonstrate the inference procedure over the
following example program which computes the mean value
of a five-point stencil:%at each index of the input array:
%%
\begin{minted}{fortran}
do i = 1, (n-1)
   do j = 1, (m-1)
      x       = a(i-1, j) + a(i+1, j)
      y       = a(i, j-1) + a(i, j+1)
      b(i, j) = (a(i, j) + x + y) / 5.0
   end do
end do
\end{minted}
%%
\subsubsection{Step 1: Standard control and data-flow analyses}
\label{sec:inf-step1}

@@

For each assignment statement whose left-hand side is an array
subscript on neighbourhood indices, a finite map is computed which
maps array variables to a set of vectors representing array
subscripts.  This finite map contains all array subscript expressions
which flow to this statement. More formally, a function
$\textsf{analyse}$ is applied to each statement in a program with the
following clause:
%
\begin{align*}
\hspace{-0.5em}
\begin{array}{lr}
\textsf{analyse}(v(\overline{e_1}) = e_2)
 := \qquad\qquad & \textit{where} \; \neigh(\overline{e_1}) \wedge \arrayTy(v)  \\[0.3em]
\multicolumn{2}{l}{\qquad \bigcup\{v' \mapsto \{
\textsf{schematic}(\bar{e})\} \mid v'(\bar{e}) \leftarrow \mathsf{flowsTo}(e_2),
  \arrayTy(v')\}}
\end{array}
\end{align*}
%
That is, we focus on assignments to an array subscript where the LHS 
indexing expression $\overline{e_1}$ is a neighbourhood index.  For
all array subscripts that flow to the right-hand side of this
statement, a finite map is constructed, mapping each array variable
to a set of indexing schemes for its subscripts, computed
with \textsf{schematic} (Defn.~\ref{def:schematic},
p.~\pageref{def:schematic}).
 Note that \textsf{schematic} is undefined if
$\bar{e}$ contains relative indices which are not neighbourhood
indices (\ie{}, these are not stencil computations we can handle).

For our example, the \textsf{analyse} function matches on
line 5, with the following set for $\textsf{flowsTo}(\texttt{a(i, j) + x +
  y})$:
%
\begin{align*}
\{\texttt{a(i-1, j)}, \texttt{a(i+1, j)}, \texttt{a(i, j-1)},
  \texttt{a(i, j+1)}, \texttt{a(i, j)}\}
\end{align*}
Subsequently the result of \textsf{analyse} on line 5 yields the map:
\begin{align*}
\texttt{a} \mapsto \{\vtwoh{-1}{0}, \vtwoh{1}{0},
          \vtwoh{0}{-1}, \vtwoh{0}{1}, \vtwoh{0}{0}\}
\end{align*}
%
If the LHS $\overline{e_1}$ contains non-$0$ neighour offsets 
then all RHS schemes are \emph{relativized} by
$\textsf{schematic}(\overline{e_1})$. For example, 
if \texttt{a(i+1) = b(i)} then relativisation of the RHS by the LHS
produces an analysis which is equivalent to the analysis of \texttt{a(i) = b(i-1)}.
